Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2019-04-24 Manuscript/manuscriptPSsuppl.tex
Small updates

  The simulation details are given in the supplementary information.

@@

\noindent
{\it POPC:POPS (5:1) mixtures with additional potassium ions} \\
The data in Table \ref{mixedIONsystems} and Refs \citenum{POPCpopsMACROG,POPCpopsMACROGwithK}

The bilayers containing a total of 120 POPC and 24 POPS molecules distributed evenly among the 
two leaflets were set up using CHARMM-GUI \cite{lee16,jo18}.
The bilayers were solvated by 5760 water molecules, corresponding to 40 water molecules for lipid.
Additionally, 24 potassium ions were added to neutralize the charge of the POPS head groups.
These bilayers were simulated using Gromacs ?? \cite{check the version} in the presence of counterions only, as well as together with different
concentrations of CaCl$_2$ or KCl. For the former, concentrations of 100~mM, 300~mM, 1~M, and 3~M 
were considered, which corresponded to the amounts of 10/20, 31/62, 104/208, and 311/622 
of Ca$^{2+}$/Cl$^-$ ions. For membranes with KCl, concentrations of 500~mM, 1~M, 2~M, and 3~M of 
were considered, corresponding to 52, 104, 208, or 311 pairs of K$^+$ and Cl$^-$ ions.

The lipids were described using the MacRog model \cite{maciejewski14,kulig15b,rog16}. The POPC topologies were obtained from 
Ref.~\citenum{kulig15c}, whereas the POPS topology was adapted from 
Ref.~\citenum{rog16}. Notably, the chains were reversed in order to modify 
OPPS into POPS. Unlike the D stereoisomer present in Ref.~\cite{rog16}, the 
POPS head group conformations generated by CHARMM-GUI are in the naturally occurring L stereoisomer.
TIP3P model \cite{jorgensen83} was used for water, and the \AA{}qvist parameters 
\cite{aqvist90}, standard for the OPLS force field, were employed for the ions.

The simulation parameters were taken from Ref.~\citenum{rog16}. 
Periodic boundary conditions were employed in all dimensions. The list of neighbors was updated every step 
\hl{I used nstlist=1 because it's used in the mdp in the article quoted. In DOI: 10.1021/jp5016627, nstlist=10 is used}
The Lennard-Jones interactions were cut off at 1~nm, whereas the smooth particle mesh Ewald algorithm 
\cite{essman95} was used for electrostatics. Dispersion correction \cite{shirts07} 
was applied to both energy and pressure. Temperature was kept constant at 298~K using the 
Nos\'{e}--Hoover thermostat \cite{nose84,hoover85}. The lipids and 
the solvent were coupled separately, and a time constant of 0.4~ps was employed. Pressure was 
maintained semi-isotropically at 1~bar using the Parrinello--Rahman barostat with a time constant of 1~ps 
\cite{parrinello81}. All non-water bonds were constrained with the LINCS algorithm 
\cite{hess97,hess07},
whereas the bonds in water molecules were constrained using SETTLE \cite{miyamoto92}.
The integration time step was set to 2~fs, and the simulations were performed either for 400~ns 
(only counterions), 300~ns (systems with KCl), or 600~ns (systems with CaCl$_2$). The last 
250, 200, or 300~ns of the simulations were included in the analyses, respectively.

@@

The structure of the bilayers with 128 DOPS or 128 POPS lipids were
obtained from CHARMM-GUI~\cite{lee16,jo18}. Bilayers are solvated with
4480 water molecules resulting in 31.1 water molecules per lipid. For
each system, 128 NaCl molecules are added to neutralize the system.\\

@@

In all simulations, TIP3P~\cite{jorgensen1983comparison} was used as the solvent. SETTLE~\cite{miyamoto92} algorithm was employed 
to constrain the bonds in TIP3P water molecules. Simulations were ran using Amber18 simulation package \cite{??} \todo{Is this correct, do we need the citation?}
and lipid parameters were obtained from Amber Lipid17 force field \cite{gould18}. For the ions, one set of simulations were run with Joung-Cheatham
ion parameters~\cite{joung2008determination} and one set of simulations were run with ff99SB~\cite{aqvist90} ion parameters. The data in Table~\ref{PSsystems} and Refs~\citenum{lipid17POPSjcions,lipid17POPSff99ions,lipid17DOPSjcions,lipid17DOPSff99ions} contain the details of the simulations and the raw trajectories, respectively.

@@

\noindent
{\it POPC:POPS (5:1) mixtures with additional calcium} \\
The data in Table \ref{mixedIONsystems} and Refs \citenum{lipid17_cacl_series,POPCpopsLIPID17withCaCl}

@@

\todo{J. Melcr, please write simulation details} \\

@@

\noindent
{\it DOPS with 288 lipids.}
The starting structure for simulation was contructed with the MEMBRANE BUILDER website~\cite{ghahremanpour13}: 288 DOPS lipids, 11232 water molecules and 288 Na ions. The Slipids~\cite{jambeck12,jambeck2012another} forcefield was used for DOPS, the TIP3P~\cite{jorgensen83} water model was used to solvate the system and Ions are described by the parameters derived by Aqvist~\cite{aqvist90}.
Simulation was performed for 200~ns, and the last 100~ns were used for the analysis. Simulation was carried out within the NPT ensemble using the GROMACS 5.0.4 package~\cite{abraham2015gromacs}. Timestep of 2~fs was used with the leapfrog integrator. The Nos\'{e}--Hoover thermostat~\cite{nose84,hoover85} was used with reference temperature of 303~K and a relaxation time constant of 0.5~ps; lipids and water plus ions were coupled separately to the heat bath. Pressure was kept constant at 1.013~bar using a semi--isotropic Parrinello--Rahman
barostat~\cite{parrinello81} with a time constant of 10.0~ps. Long-range electrostatic interactions were calculated using the PME method~\cite{darden93,essman95}. A real space cut-off of 1.0~nm was employed with grid spacing of 0.12~nm in the reciprocal space. Lennard-Jones potentials were cut off at 1.4~nm, with a dispersion correction applied to both energy and pressure. All covalent bonds in lipids were constrained using the LINCS algorithm~\cite{hess97}, whereas water molecules were constrained using SETTLE~\cite{miyamoto92}. Twin-range cutoffs, 1.0~nm and 1.6~nm, were used for the neighbor lists with the long-range neighbor list updated every 10 steps.

\todo{Other systems to be written by Piggot}

%{\it DOPS and POPS bilayers.} \\
%The data in Table \ref{PSsystems} and Refs. \citenum{slipidsPOPS298K,slipidsDOPS303K,slipidsDOPSfiles}.

@@

\noindent
{\it POPC:POPS (4:1) mixtures with additional sodium and calcium ions.} 
Previously published simulations with the additional amount of sodium \cite{jurkiewicz12}
available from Ref. \citenum{POPCpopsBERGERwith1000mMNa} and with the additional amounts
of calcium \cite{melcrova16} available from Refs. \citenum{POPCpopsBERGERwith102mMCa,POPCpopsBERGERwith715mMCa}
were used. The additional sodium ions were removed from the structure file of the simulation
with excess amount of sodium \cite{POPCpopsBERGERwith1000mMNa} to generate the initial structure for the reference simulation
with sodium counterions only. The data, input and force field parameter files of the reference simulation
are available from Ref. \citenum{bergerPOPSPOPC4:1mixtureT310K}. \todo{The simulation details of this are
  not clearly described in Zenodo repository, but can be read from the mdp file available there. Should we
write the detailes here or not?}

@@

\pagebreak
\section{Sensitivity of the molecular electrometer to the chosen definition of ion concentration}
%\section{Effect of the definition of concentration on the headgroup order parameters as a function of ion concentration.}\label{concentrationDEFsection}
Previous studies using the electrometer concept to assess the ion binding affinity to lipid
bilayers report ion concentrations either in water before solvating the lipids (buffer concentration) \cite{akutsu81,roux90,catte16}
or in bulk water after solvating the lipids (bulk concentration) \cite{altenbach84,melcr18}.
In this work, we use the former definition to be consistent with the experimental reference data \cite{roux90}. The difference between these two concentrations increases
with the increasing ion binding affinity. However, Fig. \ref{concentrationDEFfigure} demonstrates that 
in a model with realistic ion binding affinity~\cite{melcr18} the deviation between the two definitions is neglible.
\begin{figure}[]
  \centering
  \includegraphics[width=8.3cm]{../Figs/OP_ECC_POPC_DPPC_water_conc2_dppc_bulk.pdf}
  \caption{\label{concentrationDEFfigure}
    Changes of the headgroup order parameters as a function of CaCl$_2$ concentration using
    two possible definitions of ion concentration from the recent force field
    with realistic calcium binding affinity to a POPC bilayer \cite{melcr18}
    together with the experimental data \cite{akutsu81,altenbach84}. 
  }
\end{figure}

@@

Interpretation of the order parameter signs of the $\alpha$ carbon from the SDROSS experiment
is complicated by the presence of distinct order parameters for the two attached hydrogens.
As discussed in the main text, interpretation of the SDROSS curve with the help of SIMPSON
simulations gave values +0.09 and -0.02 for the $\alpha$-carbon order parameters.
To corroborate our interpretation, we measured the SDROSS curve also using the higher 8 kHz
MAS frequence (Fig. \ref{DPslices} bottom), which makes the experiment more sensitive to larger order parameter values.
Also this experiment indicates a postive value for the larger $\alpha$-carbon order parameter.
\begin{figure}[]
%  \centering
  \includegraphics[width=\textwidth]{../Figs/SI_man.pdf}
  \caption{\label{DPslices}
    Spectral slices in indirect dimension from the R-PDLF (top) and S-DROSS (middle and bottom) experiments at distinct chemical shifts.
    The chemical shifts and order parameters (calculated from the dipolar splittings indicated with blue lines in the R-PDLF slices)
    are shown on top each column (chemical shift left, order parameter right).
    The assignment of columns is given in the middle.  
    The SDROSS slices were measured using both 5 KHz (middle) and 8 kHz (bottom) because different MAS frequencies enable the
    higher sensitivity to the order parameters with lower and higher magnitudes, respectively.
    The background noise taken from chemical shift slices without carbon peaks (grey lines in SDROSS figures)
    are used to determine the error bars for $\alpha$ and $\beta$ carbons in figure \ref{PShgSIGNSsimpson} in the main text.
    The SDROSS profile of the acyl chain methyl carbons (left column) was used as reference assuming that these carbons have negative order parameters.  
  }
\end{figure}

@@

In simulations with Berger and CHARMM36 models, the responses of POPC and POPS order parameter
of to added sodium and potassium are not in line with the experiments. Instead, the simulations produces a response similar to experiments conducted in LiCl (Fig. \ref{PSresponseTONaCl}), indicating overestimated binding affinity of sodium and potassium 
in these simulations. The MacRog simulations with potassium exhibit weaker counterion binding affinity
(Fig. \ref{CIdensPSOCmixt}), but significantly larger error bars and
less systematic changes in the order parameters (Fig. \ref{PSresponseTONaCl}).
Similar unsystematic behaviour was also observed in the simulations of Lipid14/17 model
with the additional counterions \cite{POPCpopsLIPID17withKCI,POPCpopsLIPID17withK,POPCpopsLIPID17withNaCI,POPCpopsLIPID17withNa},
for which the data is not shown due to the formation of
ion clusters in water with relatively low (1 M) ion concentrations (Fig. \ref{ionCLUSTERS}).
Appearance of such clusters in the MagRog simulations with 4 M KCl
could explain the unsystematic changes of the order parameters in this model upon increasing KCl.
In conclusion, the results are in line with the section \ref{ciBINDINGsection}
in the main text, suggesting that the MacRog simulations with KCl give the most
realistic surface charge at the lipid bilayer interface among the tested simulation models.
\begin{figure*}[]
  \centering
  \includegraphics[width=17.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture at 298~K
    (except Berger simulations are (4:1) mixture at 310~K).
    The experimental data is from Ref. \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is shifted with the same value for both order parameters such that the lower order
    parameter value is at zero.
  }
\end{figure*}
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
  \caption{  Counterion density distributions from PC:PS mixtures.
\label{CIdensPSOCmixt}
  }
\end{figure}
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/lipid17cluster.png}
  \includegraphics[width=8.0cm]{../Figs/MacRogIONcluster.png}
  \caption{Ion clusters appearing in POPC:POPS (5:1) lipid17/14 simulations with 1 M of NaCl (left)
    and MacRog simulations with 4 M of KCl (right).
\label{ionCLUSTERS}
  }
\end{figure}
\newpage

@@

The response of POPC headgroup order parameters to the CaCl$_2$ concentration
are underestimated in simulations of POPC:POPS (5:1) mixture with CHARMM36 when employing NBfix (as obtained from CHARMM\_GUI in January 2018) for interactions between calcium and lipid oxygens \cite{kim16}
(Fig. \ref{changesWITHCaClPS} in the main text),
indicating that the calcium binding to the bilayer is too weak with these
parameters. The response of headgroup order
parameters (Fig. \ref{OP_CHARMM_CaCl_POPC_NBFix}) and the binding affinity
(Fig. \ref{density_profile_CHARMM_CaCl_POPC_NBFix}) of calcium also to a pure
POPC bilayer are underestimated when using NBfix ions.  Notably, CHARMM36 simulations with the NBfix
terms \cite{venable13,kim16} predict similar binding affinity for sodium and calcium. Without the NBfix term, the calcium binding affinity to pure POPC lipid bilayers was overestimated in the CHARMM36 model \cite{catte16}.

@@

\pagebreak
\section{Calcium density profiles from simulations with \\ POPC:POPS(5:1) mixture}
\todo{This section could benefit from some clarifying text}
\begin{figure}[ht]
  \centering
  \includegraphics[width=10cm]{../Figs/CAdensPCPSmixture.eps}
  \caption{\label{CAdensPCPSmixtureALL}
    Number density profiles of Ca$^{2+}$ from POPC:POPS (5:1) mixtures simulated with different force fields. The ion densities are taken along the z-axis that coincides with the bilayer normal. 
  }
  \todo{Should we include also counterions into the plot?} \\
  \todo{It would be helpful to have the panels in the same order as in the main text}
\end{figure}

\pagebreak
\section{Details of the rough subjective force field ranking (Fig.~\ref{comparisonTablePS})}

@@

\newpage
\bibliography{refs.bib}

>>>>

  % The simulation details are given in the supplementary information.

@@

\noindent
{\it POPC:POPS (5:1) mixtures with additional potassium ions} \\
%The data in Table \ref{mixedIONsystems} and Refs \citenum{POPCpopsMACROG,POPCpopsMACROGwithK}
The bilayers containing a total of 120 POPC and 24 POPS molecules distributed evenly among the 
two leaflets were set up using CHARMM-GUI \cite{lee16,jo18}.
The bilayers were solvated by 5760 water molecules, corresponding to 40 water molecules for lipid.
Additionally, 24 potassium ions were added to neutralize the charge of the POPS head groups.
These bilayers were simulated using Gromacs 2016.3~\cite{abraham2015gromacs} in the presence of counterions only, as well as together with different
concentrations of CaCl$_2$ or KCl. For the former, concentrations of 100~mM, 300~mM, 1~M, and 3~M 
were considered, which corresponded to the amounts of 10/20, 31/62, 104/208, and 311/622 
of Ca$^{2+}$/Cl$^-$ ions. For membranes with KCl, concentrations of 500~mM, 1~M, 2~M, and 3~M of 
were considered, corresponding to 52, 104, 208, or 311 pairs of K$^+$ and Cl$^-$ ions.

The lipids were described using the MacRog model \cite{maciejewski14,kulig15b,rog16}. The POPC topologies were obtained from 
Ref.~\citenum{kulig15c}, whereas the POPS topology was adapted from OPPS topology in
Ref.~\citenum{rog16} by reversing the order of acyl chains.
The initial structures from CHARMM-GUI were used because they have the
naturally occurring L stereoisomer in the POPS head group, instead of the 
the D stereoisomer present in Ref.~\citenum{rog16}.
TIP3P model \cite{jorgensen83} was used for water, and the \AA{}qvist parameters 
\cite{aqvist90}, standard for the OPLS force field, were employed for the ions.

The simulation parameters were taken from Ref.~\citenum{rog16}. 
Periodic boundary conditions were employed in all dimensions. The list of neighbors was updated every step. 
%\hl{I used nstlist=1 because it's used in the mdp in the article quoted. In DOI: 10.1021/jp5016627, nstlist=10 is used}
The Lennard-Jones interactions were cut off at 1~nm, whereas the smooth particle mesh Ewald algorithm 
\cite{essman95} was used for electrostatics. Dispersion correction \cite{shirts07} 
was applied to both energy and pressure. Temperature was kept constant at 298~K using the 
Nos\'{e}--Hoover thermostat \cite{nose84,hoover85}. The lipids and 
the solvent were coupled separately, and a time constant of 0.4~ps was employed. Pressure was 
maintained semi-isotropically at 1~bar using the Parrinello--Rahman barostat with a time constant of 1~ps 
\cite{parrinello81}. All non-water bonds were constrained with the LINCS algorithm 
\cite{hess97,hess07},
whereas the bonds in water molecules were constrained using SETTLE \cite{miyamoto92}.
The integration time step was set to 2~fs, and the simulations were performed either for 400~ns 
(only counterions), 300~ns (systems with KCl), or 600~ns (systems with CaCl$_2$). The last 
250, 200, or 300~ns of the simulations were included in the analyses, respectively.

@@

The structure of the bilayers with 128 DOPS or 128 POPS lipids were
obtained from CHARMM-GUI~\cite{lee16,jo18}. Bilayers are solvated with
4480 water molecules resulting in 31.1 water molecules per lipid. For
each system, 128 NaCl molecules are added to neutralize the system.
In all simulations, TIP3P~\cite{jorgensen1983comparison} was used as the solvent. SETTLE~\cite{miyamoto92} algorithm was employed 
to constrain the bonds in TIP3P water molecules. Simulations were ran using Amber18 simulation package \cite{??} \todo{Is this correct, do we need the citation?}
and lipid parameters were obtained from Amber Lipid17 force field \cite{gould18}. For the ions, one set of simulations were run with Joung-Cheatham
ion parameters~\cite{joung2008determination} and one set of simulations were run with ff99SB (\AA{}qvist)~\cite{aqvist90} ion parameters.
The data in Table~\ref{PSsystems} and Refs~\citenum{lipid17POPSjcions,lipid17POPSff99ions,lipid17DOPSjcions,lipid17DOPSff99ions} contain
the number of molecules in simulations, related simulation files and trajectories.

@@

@@

\noindent
{\it POPC:POPS (5:1) mixtures with additional calcium} \\
%The data in Table \ref{mixedIONsystems} and Refs \citenum{lipid17_cacl_series,POPCpopsLIPID17withCaCl}

@@

\todo{J. Melcr, please write simulation details about Gromacs simulations with Dang ions.} \\

@@

\noindent
{\it POPS}
\todo{POPS \cite{slipidsPOPS298K} to be written by Piggot}

\noindent
{\it DOPS with 128 lipids.}
\todo{DOPS with 128 lipids \cite{slipidsDOPS303K} to be written by Piggot}

\noindent
{\it DOPS with 288 lipids.}
The starting structure with 288 DOPS lipids, 11232 water molecules and 288 Na ions
was contructed with the MEMBRANE BUILDER website~\cite{ghahremanpour13}.
The Slipids~\cite{jambeck12,jambeck2012another} forcefield was used for DOPS, the TIP3P~\cite{jorgensen83} water model was used to solvate the system and
ions are described by the parameters derived by \AA{}qvist~\cite{aqvist90}.
The system was simulated in NPT ensemble for 200~ns using the GROMACS 5.0.4 package~\cite{abraham2015gromacs},
and the last 100~ns was used for the analysis.
Timestep of 2~fs was used with the leapfrog integrator. The Nos\'{e}--Hoover thermostat~\cite{nose84,hoover85} was used with reference temperature of 303~K and a relaxation time constant of 0.5~ps; lipids and water plus ions were coupled separately to the heat bath. Pressure was kept constant at 1.013~bar using a semi--isotropic Parrinello--Rahman
barostat~\cite{parrinello81} with a time constant of 10.0~ps. Long-range electrostatic interactions were calculated using the PME method~\cite{darden93,essman95}. A real space cut-off of 1.0~nm was employed with grid spacing of 0.12~nm in the reciprocal space. Lennard-Jones potentials were cut off at 1.4~nm, with a dispersion correction applied to both energy and pressure. All covalent bonds in lipids were constrained using the LINCS algorithm~\cite{hess97}, whereas water molecules were constrained using SETTLE~\cite{miyamoto92}. Twin-range cutoffs, 1.0~nm and 1.6~nm, were used for the neighbor lists with the long-range neighbor list updated every 10 steps.

@@

\noindent
{\it POPC:POPS (4:1) mixtures with additional sodium and calcium ions.} 
Previously published simulations with the additional amount of sodium \cite{jurkiewicz12}
available from Ref. \citenum{POPCpopsBERGERwith1000mMNa} and with the additional amounts
of calcium \cite{melcrova16} available from Refs. \citenum{POPCpopsBERGERwith102mMCa,POPCpopsBERGERwith715mMCa}
were used. To generate the reference system with sodium counterions only,
the additional ions were removed from the structure file in Ref.~\citenum{POPCpopsBERGERwith1000mMNa}.
This system was simulated 120~ns using Gromacs 5~\cite{abraham2015gromacs} with parameter (mdp file)
and force field files (top and itp) available from Ref.~\citenum{bergerPOPSPOPC4:1mixtureT310K},
the last 80~ns was used for the analysis.
%\todo{The simulation details of this are not clearly described in Zenodo repository, but can be read from the mdp file available there.
%  Should we write the detailes here or not?}

@@

\pagebreak
\section{Sensitivity of the molecular electrometer to the chosen definition of ion concentration}\label{concentrationDEFsection}
%\section{Effect of the definition of concentration on the headgroup order parameters as a function of ion concentration.}\label{concentrationDEFsection}
Previous studies using the electrometer concept to assess the ion binding affinity to lipid
bilayers report ion concentrations either in water before solvating the lipids (buffer concentration) \cite{akutsu81,roux90,catte16}
or in bulk water after solvating the lipids (bulk concentration) \cite{altenbach84,melcr18}.
In this work, we use the former definition to be consistent with the experimental reference data \cite{roux90}. The difference between these two concentrations increases
with the increasing ion binding affinity. However, Fig. \ref{concentrationDEFfigure} demonstrates that 
in a model with realistic ion binding affinity~\cite{melcr18} the deviation between the two definitions is neglible.
\begin{figure}[]
  \centering
  \includegraphics[width=8.3cm]{../Figs/OP_ECC_POPC_DPPC_water_conc2_dppc_bulk.pdf}
  \caption{\label{concentrationDEFfigure}
    Changes of the headgroup order parameters as a function of CaCl$_2$ concentration using
    two possible definitions of ion concentration from the recent force field
    with realistic calcium binding affinity to a POPC bilayer \cite{melcr18}
    together with the experimental data \cite{akutsu81,altenbach84}. 
  }
\end{figure}

@@

Interpretation of the order parameter signs of the $\alpha$ carbon from the SDROSS experiment
is complicated by the presence of distinct order parameters for the two attached hydrogens.
As discussed in the main text, interpretation of the SDROSS curve with the help of SIMPSON
simulations gave values +0.09 and -0.02 for the $\alpha$-carbon order parameters.
To corroborate our interpretation, we measured the SDROSS curve also using the higher 8 kHz
MAS frequence (Fig. \ref{DPslices} bottom), which makes the experiment more sensitive to larger order parameter values.
Also this experiment indicates a positive value for the larger $\alpha$-carbon order parameter.
\begin{figure}[]
%  \centering
  \includegraphics[width=\textwidth]{../Figs/SI_man.pdf}
  \caption{\label{DPslices}
    Spectral slices in indirect dimension from the R-PDLF (top) and S-DROSS (middle and bottom) experiments at distinct chemical shifts.
    The chemical shifts and order parameters (calculated from the dipolar splittings indicated with blue lines in the R-PDLF slices)
    are shown on top each column (chemical shift left, order parameter right).
    The assignment of columns is given in the middle.  
    The SDROSS slices were measured using both 5 KHz (middle) and 8 kHz (bottom) because different MAS frequencies enable the
    higher sensitivity to the order parameters with lower and higher magnitudes, respectively.
    The background noise taken from chemical shift slices without carbon peaks (grey lines in SDROSS figures)
    are used to determine the error bars for $\alpha$ and $\beta$ carbons in figure \ref{PShgSIGNSsimpson} in the main text.
    The SDROSS profile of the acyl chain methyl carbons (left column) was used as reference assuming that these carbons have negative order parameters.  
  }
\end{figure}

@@

In simulations with Berger and CHARMM36 models, the responses of POPC and POPS order parameter
to added sodium and potassium are not in line with the experiments. Instead, the simulations produces a response similar to experiments conducted in LiCl (Fig. \ref{PSresponseTONaCl}), indicating overestimated binding affinity of sodium and potassium 
in these simulations. The MacRog simulations with potassium exhibit weaker counterion binding affinity
(Fig. \ref{CIdensPSOCmixt}), but significantly larger error bars and
less systematic changes in the order parameters (Fig. \ref{PSresponseTONaCl}).
Similar unsystematic behaviour was also observed in the simulations of Lipid14/17 model
with the additional counterions \cite{POPCpopsLIPID17withKCI,POPCpopsLIPID17withK,POPCpopsLIPID17withNaCI,POPCpopsLIPID17withNa},
for which the data is not shown due to the formation of
ion clusters in water with relatively low (1 M) ion concentrations (Fig. \ref{ionCLUSTERS}).
Appearance of such clusters in the MagRog simulations with 4 M KCl
could explain the unsystematic changes of the order parameters in this model upon increasing KCl.
In conclusion, the results are in line with the section \ref{ciBINDINGsection}
in the main text, suggesting that the MacRog simulations with KCl give the most
realistic surface charge at the lipid bilayer interface among the tested simulation models.
\begin{figure*}[]
  \centering
  \includegraphics[width=17.0cm]{../Figs/CHANGESwithMONVALENTwithPS.eps}
  \caption{\label{PSresponseTONaCl}
    Changes of the PC (left) and PS (right) headgroup order parameters as a function of
    added NaCl, KCl and LiCl from POPC:POPS (5:1) mixture at 298~K
    (except Berger simulations are (4:1) mixture at 310~K).
    The experimental data is from Ref. \citenum{roux90}.
    The values from counterion-only systems are set as a zero point of y-axis.
    To correctly illustrate the significant forking of the $\alpha$-carbon order parameter
    in PS headgroup (bottom, right), the y-axis is shifted with the same value for both order parameters such that the lower order
    parameter value is at zero.
  }
\end{figure*}
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/CIdensPSOCmixt.eps}
  \caption{  Counterion density distributions from PC:PS mixtures.
\label{CIdensPSOCmixt}
  }
\end{figure}
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/lipid17cluster.png}
  \includegraphics[width=8.0cm]{../Figs/MacRogIONcluster.png}
  \caption{Ion clusters appearing in POPC:POPS (5:1) lipid17/14 simulations with 1 M of NaCl (left)
    and MacRog simulations with 4 M of KCl (right).
\label{ionCLUSTERS}
  }
\end{figure}
\newpage

@@

The response of POPC headgroup order parameters to the CaCl$_2$ concentration
are underestimated in simulations of POPC:POPS (5:1) mixture with CHARMM36 when employing NBfix (as obtained from CHARMM-GUI in January 2018) for interactions between calcium and lipid oxygens \cite{kim16}
(Fig. \ref{changesWITHCaClPS} in the main text),
indicating that the calcium binding to the bilayer is too weak with these
parameters. The response of headgroup order
parameters (Fig. \ref{OP_CHARMM_CaCl_POPC_NBFix}) and the binding affinity
(Fig. \ref{density_profile_CHARMM_CaCl_POPC_NBFix}) of calcium also to a pure
POPC bilayer are underestimated when using NBfix ions.  Notably, CHARMM36 simulations with the NBfix
terms \cite{venable13,kim16} predict similar binding affinity for sodium and calcium. Without the NBfix term, the calcium binding affinity to pure POPC lipid bilayers was overestimated in the CHARMM36 model \cite{catte16}.

@@

\clearpage
\section{Calcium density profiles from simulations with \\ POPC:POPS(5:1) mixture}
%\todo{This section could benefit from some clarifying text}
\begin{figure}[ht]
  \centering
  \includegraphics[width=10cm]{../Figs/CAdensPCPSmixture.eps}
  \caption{\label{CAdensPCPSmixtureALL}
    Number density profiles of Ca$^{2+}$ from POPC:POPS (5:1) mixtures simulated with different force fields. The ion densities are taken along the z-axis that coincides with the bilayer normal. 
  }
%  \todo{Should we include also counterions into the plot?} \\
\end{figure}

\clearpage
\section{Details of the rough subjective force field ranking (Fig.~\ref{comparisonTablePS})}

@@

\section{Author contributions}
\noindent
{\it Hanne Antila}

\noindent
{\it Pavel Buslaev}

\noindent
{\it Fernando Favela}

\noindent
{\it Tiago M. Ferreira}

\noindent
{\it Ivan Gushchin}

\noindent
{\it Matti Javanainen}

\noindent
{\it Batuhan Kav}

\noindent
{\it Jesper J. Madsen}

\noindent
{\it Josef Melcr}

\noindent
{\it Markus Miettinen}

\noindent
{\it Ricky Nencini}

\noindent
{\it O. H. Samuli Ollila}
Designed the project and managed the work.
Ran and analysed several simulations. Wrote the manuscript.

\noindent
{\it Thomas Piggot}

\clearpage
\bibliography{refs.bib}
